---
title: "Lab_10"
subtitle: "ESS-330-Lab"
author: 
- name: "Aldair Perez Marmolejo"
email: "alpema@colostate.edu"
format: html
execute: 
  echo: true
---

```{r}
library(tidyverse)
library(sf)
library(units)

library(AOI)

library(gghighlight)
library(ggrepel)
library(knitr)

library(flextable)
library(ggplot2)
library(dplyr)
library(tibble)
```

# Question 1
```{r}
eqdc <- '+proj=eqdc +lat_0=40 +lon_0=-96 +lat_1=20 +lat_2=60 +x_0=0 +y_0=0 +datum=NAD83 +units=m +no_defs'

remotes::install_github("mikejohnson51/AOI", force = TRUE)
```

```{r}
Pais <- aoi_get(country = c("MX", "CA", "USA"))
pais_eqdc <- st_transform(Pais, crs = eqdc)
st_crs(pais_eqdc)
```

```{r}
url <- "https://simplemaps.com/static/data/us-cities/1.90/basic/simplemaps_uscities_basicv1.90.zip"
zip <- tempfile(fileext = ".zip")
download.file(url, zip)
unzip(zip, exdir = tempdir())
csv_file <- list.files(tempdir(), pattern = "uscities.csv$", full.names = TRUE)
cities_df <- read_csv(csv_file)
```

```{r}
cities_sf <- st_as_sf(cities_df, coords = c("lng", "lat"), crs = 4326)
cities_sf$state_id <- as.character(cities_sf$state_id)
excluded_states <- c("AK", "HI", "PR")
cities_filtered <- cities_sf %>%
  filter(!state_id %in% excluded_states)
```

# Question 2
## 2.1: Border
```{r}
usa_poly <- Pais %>%
  st_union() %>%
  st_make_valid()
usa_border <- st_cast(usa_poly, "MULTILINESTRING")

cities_proj <- st_transform(cities_filtered, crs = st_crs(pais_eqdc))
usa_border_proj <- st_transform(usa_border, crs = st_crs(pais_eqdc))

cities_proj$distance_to_usa_border <- st_distance(cities_proj, usa_border_proj) / 1000

cities_farthest_usa_border <- cities_proj %>%
  st_drop_geometry() %>%
  arrange(desc(distance_to_usa_border)) %>%
  head(5)

flextable(cities_farthest_usa_border[, c("city", "state_id", "distance_to_usa_border")])
```

## 2.2: States
```{r}
usa_states <- Pais %>% 
  st_cast("MULTILINESTRING") %>%
  st_union() %>%
  st_make_valid()

usa_states_proj <- st_transform(usa_states, crs = st_crs(pais_eqdc))

cities_proj$distance_to_state_border <- st_distance(cities_proj, usa_states_proj) / 1000

cities_farthest_state_border <- cities_proj %>%
  st_drop_geometry() %>%
  arrange(desc(distance_to_state_border)) %>%
  head(5)

flextable(cities_farthest_state_border[, c("city", "state_id", "distance_to_state_border")])
```

## 2.3: Mexico
```{r}
mexico_border_proj <- Pais %>%
  filter(adm0_a3 == "MEX") %>%
  st_union() %>%
  st_make_valid() %>%
  st_transform(crs = st_crs(pais_eqdc)) %>%
  st_cast("MULTILINESTRING")

cities_proj$distance_to_mexico_border <- as.numeric(st_distance(cities_proj, mexico_border_proj)) / 1000

cities_farthest_mexico_border <- cities_proj %>%
  st_drop_geometry() %>%
  arrange(desc(distance_to_mexico_border)) %>%
  head(5)

flextable(cities_farthest_mexico_border[, c("city", "state_id", "distance_to_mexico_border")])
```

## 2.4: Canada
```{r}
canada_border <- Pais %>% 
  filter(adm0_a3 == "CAN") %>%
  st_union() %>%
  st_make_valid()
canada_border <- st_cast(canada_border, "MULTILINESTRING")

canada_border_proj <- st_transform(canada_border, crs = st_crs(pais_eqdc) )

cities_proj$distance_to_canada_border <- st_distance(cities_proj, canada_border_proj) / 1000

cities_farthest_canada_border <- cities_proj %>%
  st_drop_geometry() %>%
  arrange(desc(distance_to_canada_border)) %>%
  head(5)

flextable(cities_farthest_canada_border[, c("city", "state_id", "distance_to_canada_border")])
```

# Question 3
```{r}
top10_cities <- cities_proj %>%
  arrange(desc(population)) %>%
  slice(1:10)
ggplot() +
  geom_sf(data = pais_eqdc, fill = "blue", color = "darkorange", lty = 1, size = 0.5) +
  geom_sf(data = Pais %>% st_transform(crs = st_crs(pais_eqdc)), fill = NA, color = "green", lty = 2, size = 0.3) +
  geom_sf(data = top10_cities, color = "red", size = 2) +
  ggrepel::geom_label_repel(data = top10_cities, aes(geometry = geometry, label = city), stat = "sf_coordinates", size = 3) +
  theme_minimal() +
  ggtitle("Top 10 USA Cities")
```

## 3.2: Border
```{r}
farthest_usa_border <- cities_proj %>%
  arrange(desc(distance_to_usa_border)) %>%
  slice(1:5)

cities_proj$distance_km <- as.numeric(cities_proj$distance_to_usa_border)

ggplot() +
  geom_sf(data = pais_eqdc, fill = "grey90", color = "black", lty = 1) +
  geom_sf(data = cities_proj, aes(color = distance_km), size = 1) +
  scale_color_viridis_c(name = "Distance to National Border (km)") +
  geom_sf(data = farthest_usa_border, color = "red", size = 2) +
  ggrepel::geom_label_repel(data = farthest_usa_border,
                            aes(geometry = geometry, label = city),
                            stat = "sf_coordinates", size = 3) +
  theme_minimal() +
  ggtitle("Distance of USA Cities to National Border")
```

## 3.3: State Border
```{r}
farthest_state_border <- cities_proj %>%
  arrange(desc(distance_to_state_border)) %>%
  slice(1:5)

ggplot() +
  geom_sf(data = pais_eqdc, fill = "grey", color = "black") +
  geom_sf(data = cities_proj, aes(color = distance_km), size = 1) +
  scale_color_viridis_c(name = "Distance to State Border (km)") +
  geom_sf(data = farthest_state_border, color = "blue", size = 2) +
  ggrepel::geom_label_repel(data = farthest_state_border,
                            aes(geometry = geometry, label = city),
                            stat = "sf_coordinates", size = 3) +
  theme_minimal() +
  ggtitle("Distance of USA Cities to Nearest State Border")
```

## 3.4: Mexico and Canada
```{r}
cities_proj$border_diff <- abs(
  as.numeric(cities_proj$distance_to_mexico_border) -
  as.numeric(cities_proj$distance_to_canada_border)
)

equidistant_cities <- cities_proj %>%
  filter(border_diff <= 100)

top5_equidistant <- equidistant_cities %>%
  arrange(desc(population)) %>%
  slice(1:5)

ggplot() +
  geom_sf(data = pais_eqdc, fill = "white", color = "black") +
  geom_sf(data = cities_proj, color = "orange", size = 0.5) +
  geom_sf(data = equidistant_cities, aes(color = border_diff), size = 1) +
  scale_color_viridis_c(name = "|Distance Mexico - Canada| (km)") +
  gghighlight(border_diff <= 100) +
  ggrepel::geom_label_repel(data = top5_equidistant,
                            aes(geometry = geometry, label = city),
                            stat = "sf_coordinates", size = 3, color = "darkred") +
  theme_minimal() +
  ggtitle("Cities Equidistant from Canada and Mexico (±100 km)")
```

# Question 4
## 4.1: Quantifying Border
```{r}
border_km <- 160

cities_proj$distance_to_usa_border <- as.numeric(cities_proj$distance_to_usa_border)

cities_border_zone <- cities_proj %>%
  filter(distance_to_usa_border <= border_km)

total_cities <- nrow(cities_proj)

total_population <- sum(cities_proj$population, na.rm = TRUE)
border_zone_population <- sum(cities_border_zone$population, na.rm = TRUE)

percent_cities_in_zone <- nrow(cities_border_zone) / total_cities * 100
percent_population_in_zone <- border_zone_population / total_population * 100

border_summary <- tibble(
  Metric = c("Total Cities", "Cities within 100 miles", "Total Population", "Population in 100-mile Zone", "Percent of Cities", "Percent of Population"),
  Value = c(total_cities,
            nrow(cities_border_zone),
            formatC(total_population, format = "d", big.mark = ","),
            formatC(border_zone_population, format = "d", big.mark = ","),
            sprintf("%.2f%%", percent_cities_in_zone),
            sprintf("%.2f%%", percent_population_in_zone))
)

flextable(border_summary)
```
### The results from the table somewhat reflect/match to the ACLU estimates from the link provided.

## 4.2: Map Border
```{r}
top10_border_zone <- cities_border_zone %>%
  arrange(desc(population)) %>%
  slice(1:10)

ggplot() +
  geom_sf(data = pais_eqdc, fill = "red", color = "orange") +
  geom_sf(data = cities_proj, color = "yellow", size = 0.3) +
  geom_sf(data = cities_border_zone, aes(color = population), size = 1) +
  scale_color_gradient(low = "orange", high = "darkred", name = "Population") +
  gghighlight(distance_to_usa_border <= border_km) +
  ggrepel::geom_label_repel(data = top10_border_zone,
                            aes(geometry = geometry, label = city),
                            stat = "sf_coordinates", size = 3, color = "violet") +
  ggtitle("Cities within 100-Mile Border Zone") +
  theme_minimal()
```

## 4.3
### The most populous city in each state within the Danger Zone are Seattle, San Fransico, Los Angeles, Dan Diego, Houston, Miami, Washington D.C, Philidelphia, New York, and Boston.